### 2021-01-29 Draft.tex
clean up

\documentclass{article}
%\usepackage[T1]{fontenc}
%\usepackage[utf8]{inputenc}
\usepackage{enumerate}
\usepackage{multirow}
\usepackage[bottom,flushmargin,hang,multiple]{footmisc}
\usepackage{etoolbox}
\makeatletter
\patchcmd\maketitle{\hb@xt@1.8em}{\hbox}{}{}
\makeatother
\patchcmd{\maketitle}{\@fnsymbol}{\@arabic}{}{}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage[margin=1.5in]{geometry}
\usepackage{float}
\usepackage{comment}
\usepackage{pgf}
\linespread{1.5}

\usepackage{tikz}
\usepackage{subcaption}

\usepackage[T1]{fontenc}
\usepackage[sfdefault,scaled=.95,light]{FiraSans}
\usepackage{newtxsf}

%\usepackage{url}
%\usepackage[square,sort,comma,numbers]{natbib}

\parindent=0pt
\usepackage{etoolbox}
\AtBeginEnvironment{quote}{\small}

\newcommand{\Vx}{\textbf{Var}_x}
\newcommand{\Ex}{\textbf{E}_x}
\newcommand{\Vz}{\textbf{Var}_z}
\newcommand{\Ez}{\textbf{E}_z}

% \renewcommand{\familydefault}{lmodern}
\begin{document}

\title{Moment matching in latent variable model}

%\subtitle{Poste de professeure, professeur en statstique et science des données}

%\titlerunning{Short form of title}        % if too long for running head

\author{C\'edric Beaulac}

\date{\today}
% The correct dates will be entered by the editor

\maketitle

\section{Intro}

\section{Probability identities}

The idea is to build two different estimators of the same value. One uses observed data and the other uses the proposed generative model.

\bigskip

The Law of Total Variance :

\begin{align}
\Vx(X) = \Ez[\Vx(X|Z)] + \Vz[\Ex(X|Z)]
\end{align}

Notice the second term is equal to :

\begin{align}
\Vz[\Ex(X|Z)] &= \Ez[\Ex(X|Z)^2] - (\Ez[(\Ex(X|Z)])^2 \\
&= \Ez[\Ex(X|Z)^2] - (\Ex[X])^2
\end{align}

Now we can combine and re-organize both equations :

\begin{align}
\Vx(X)+ (\Ex[X])^2 = \Ez[\Vx(X|Z)] + \Ez[\Ex(X|Z)^2]
\label{rhslhs}
\end{align}

We have re-organized both term in this particular way for one reason, the left hand side of equation \ref{rhslhs} is independent of the latent variables and of the choice of model while the right hand side contains information about both components of the generative model. Additionally, the left hand side is actually $\Ex[X^2]$, the second moment of $X$ and thus our work here consist of comparing two different estimator of $\Ex[X^2]$. The left hand side can be estimated using the observed data:

\begin{align}
\Ex[X^2] = \Vx(X)+ (\Ex[X])^2 \approx \frac{\sum_{i=1}^n(X_i-\bar{X})^T(X_i-\bar{X})}{n-1} + \bar{X}^T\bar{X} = \frac{\sum_{i=1}^n X_i^TX_i}{n-1}
\label{lhs}
\end{align}

We call the LHS the data estimate (DE). The right and side of equation \ref{rhslhs} can be estimated using every components of the proposed generative model; we estimate it via a Monte Carlo sample of the prior $p_\theta(z)$ and both $\Vx(X|Z)$ and $\Ex(X|Z)$ which completely defined $p_\theta(x|z)$ contribute to this term:

\begin{align}
\Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
&\approx \frac{1}{m} \sum_{i=1}^m ( \text{diag}(\text{MLP}_2(z_i)) + \text{MLP}_1(z_i)^T\text{MLP}_1(z_i))
\label{rhs}
\end{align}

where $z_i \sim p_\theta(z)$, MLP$_1$ is the multilayer perceptron that return $\Ex(X|Z)$ and MLP$_2$ is the multilayer perceptron that return $\Vx(X|Z)$. This is the forward model estimates (FME).

\bigskip

The estimates from equation \ref{lhs} should be similar to those from equation \ref{rhs}. Right now, we take the element-wise ratio of both these estimates and if the propose generative models does capture well the second moment of $X$ we expect elements of this matrix to be close to one.

\bigskip

Based on simulation we observe that the generative model tends to underestimate the second moment.

\bigskip

Know that we can do something similar for the first moment :

\begin{align}
\Ex[X] = \Ez[\Ex(X|Z)] 
\label{rhs}
\end{align}

where we can estimate the left-hand side with $\bar{x}$ and the right-hand side with $\frac{1}{m} \sum_{i=1}^m \text{MLP}_1(z_i)$ where $z_i \sim p_\theta(z)$. However, based on our previous experiment we do not believe capturing the first moment is an issue vor VAE at the current moment and the RHS estimate only account for a subset of the estimated parameters (it does not include the parameters of MLP$_2$).

\bigskip

The first objective is to use such strategies as diagnostic tools to evaluate the fit of latent variable model. A follow-up project is to use those as part of the optimization procedure.

\subsection{Moment matching for diagnostic}

For latent variable with a define forward mechanism, such as pPCA, GMM, HMM and VAE we can assess the ability of the forward model to capture the moment of the observation by comparing the DE with the FME.

\pagebreak
\bibliographystyle{plain}
\bibliography{mybibfile}
\bigskip

\end{document}

### 2021-01-29 NNutility.py
GMM Generator + Basic VAE model

@@ -2,7 +2,6 @@
 """
 Created on 22/01/21
 
-
 @author: Cedric Beaulac
 Clean place to drop NN Related stuff for LVS experiments
 """
@@ -18,115 +17,75 @@ import numpy as np
 from numpy import *
 import pandas as pd
 
-def train(args, model, device, dataset, optimizer, epoch,yindex):
+
+####################################
+# Defining our VAE
+####################################
+
+class PVAE(nn.Module):
+    def __init__(self, x_dim=10,h_dim=400,z_dim=20):
+        super(PVAE, self).__init__()
+
+        self.fc1 = nn.Linear(x_dim, h_dim)
+        self.fc21 = nn.Linear(h_dim, z_dim)
+        self.fc22 = nn.Linear(h_dim, z_dim)
+        self.fc3 = nn.Linear(z_dim, h_dim)
+        self.fc41 = nn.Linear(h_dim, x_dim)
+        self.fc42 = nn.Linear(h_dim, x_dim)
+
+    def encode(self, x):
+        h1 = F.relu(self.fc1(x))
+        return self.fc21(h1), self.fc22(h1)
+
+    def reparameterize(self, mu, logvar):
+        std = torch.exp(0.5*logvar)
+        eps = torch.randn_like(std)
+        return mu + eps*std
+
+    def decode(self, z):
+        h3 = F.relu(self.fc3(z))
+        return self.fc41(h3),self.fc42(h3)
+
+    def forward(self, x):
+        mu, logvar = self.encode(x)
+        z = self.reparameterize(mu, logvar)
+        mux, logvarx = self.decode(z)
+        return mux,logvarx, mu, logvar
+    
+    
+    
+####################################
+# Define the training procedure
+####################################
+# Reconstruction + KL divergence losses summed over all elements and batch
+def ploss_function(mux,logvarx, x, muz, logvarz,beta,device):
+    
+    sigmax = logvarx.mul(0.5).exp_()
+    #sigmax = torch.ones(logvarx.shape)
+    #sigmax = sigmax.to(device)
+    pxn = torch.distributions.normal.Normal(mux, sigmax)
+    logpx = torch.sum(pxn.log_prob(x))
+
+
+    KLD = 0.5 * torch.sum(1 + logvarz - muz.pow(2) - logvarz.exp())
+
+
+    return logpx + beta*KLD
+
+def ptrain(args, model, device, data, optimizer, epoch):
     model.train()
[…]

### 2021-01-29 main.py

@@ -2,12 +2,9 @@
 """
 Created on 22/01/21
 
-
-
 @author: Cedric Beaulac 
-Thank to https://github.com/pytorch/examples/blob/master/mnist/main.py
 
-First test with VAE for HWD-BR data set
+LVM-MEGA
 """
 
 
@@ -56,10 +53,168 @@ import matplotlib.pyplot as plt
 ####################################
 # Import my libraries
 ####################################
-os.chdir("C:/Users/beaul/Dropbox/U of T/Year 5/ComputerVisionInference/Python_Project4")  
 
 from utility import *
-from VAEutility import *
+from NNutility import *
 
 
-n=10pi=[0.2,0.8],mu=[0,100],std=[1,1]
+
+####################################
+# Generate data sets
+####################################
+
+
+n=5000
+pi=[0.3,0.3,0.4]
+mu=[[0,10],[-5,0],[5,2]]
+std=[[1,2],[0.5,0.5],[2,2]]
+
+sample = GMM_Gen2(n,pi,mu,std)
+
+plt.scatter(sample[:,0].numpy(),sample[:,1].numpy())
+
+
+###########################################
+# Fixing arguments
+###########################################
+
+
+ObsDim = shape(mu)[1]
+BSize = 500
+NB = int(n/BSize)
+
+parser = argparse.ArgumentParser(description='LVM MEGA')
+parser.add_argument('--batch-size', type=int, default=BSize, metavar='N',
+                        help='input batch size for training (default: 64)')
+parser.add_argument('--Number-batch', type=int, default=NB, metavar='N',
+                        help='input batch size for testing (default: 1000)')
+parser.add_argument('--test-batch-size', type=int, default=1000, metavar='N',
+                        help='input batch size for testing (default: 1000)')
+parser.add_argument('--epochs', type=int, default=15, metavar='N',
+                        help='number of epochs to train (default: 14)')
+parser.add_argument('--lr', type=float, default=1.0, metavar='LR',
+                        help='learning rate (default: 1.0)')
+parser.add_argument('--gamma', type=float, default=0.9, metavar='M',
+                        help='Learning rate step gamma (default: 0.7)')
+parser.add_argument('--no-cuda', action='store_true', default=False,
+                        help='disables CUDA training')
[…]

### 2021-01-29 utility.py

@@ -6,6 +6,7 @@ Created on 22/01/21
 Clean place to drop function for LVS experiments
 """
 
+import torch
 ##################################  
 # GMM Generator
 ##################################

### 2021-01-30 Draft.tex
Implemented Frob and pimped out draft

@@
 
-\title{Moment matching in latent variable model}
+\title{Studying the gap between moment estimators to evaluate latent variable models}
 
@@
 
-\section{Probability identities}
+\section{Latent variable models}
+
+\section{Moment estimators}
+
+\subsection{Second moment}
 
@@
 
-\bigskip
+\subsection{First moment}
 
@@
 
-\subsection{Moment matching for diagnostic}
+\section{Gap between moment estimators}
 
 For latent variable with a define forward mechanism, such as pPCA, GMM, HMM and VAE we can assess the ability of the forward model to capture the moment of the observation by comparing the DE with the FME.
+From now we refer to the object we are studying as MEGA (Moment Estimators GAp).
+
+\subsection{Selecting a matrix norm}
+
+Finally, in order to map the MEGA to something tangible and comparable we propose using a matrix norm of the MEGA.
+There exist a wide range of possible candidate, let us introduce a few a justify our finale choice.
+
+\bigskip
+
+We want to use a norm that looks at the global properties of the matrix and fortunately Rigollet \cite{rigollet15} introduces multiple norms that do so.
+
+\subsection{Implementation of the selected norms}
+
+\section{Experiments and results}
+
+\section{Training with moment estimator matching}
+
+\section{Conclusion}
 

### 2021-01-30 NNutility.py

@@ -19,9 +19,14 @@ import pandas as pd
 
 
 ####################################
-# Defining our VAE
+# Defining VAEs
 ####################################
 
+####################################
+# PVAE: Probabilistic VAE 
+# Encoder has 1 hidden layer
+# Decoder has one hidden layer
+####################################
 class PVAE(nn.Module):
     def __init__(self, x_dim=10,h_dim=400,z_dim=20):
         super(PVAE, self).__init__()
@@ -86,6 +91,13 @@ def ptrain(args, model, device, data, optimizer, epoch):
         loss.backward()
         optimizer.step()
 
-    train_loss /= data.shape[0]         
-    print('====> Epoch: {} Average loss: {:.4f}'.format(
-          epoch, train_loss ))
+    #train_loss /= data.shape[0]         
+    #print('====> Epoch: {} Average loss: {:.4f}'.format(
+    #      epoch, train_loss ))
+    
+def ptest(args, model, device, data, epoch):
+    mux,logvarx, mu, logvar = model(data)
+    loss = -ploss_function(mux,logvarx, data, mu, logvar,args.beta,device)
+     
+    print('====> Epoch: {} Average test loss: {:.4f}'.format(
+          epoch, loss/data.shape[0] ))

### 2021-01-30 main_GMM.py

@@ -4,7 +4,7 @@ Created on 22/01/21
 
 @author: Cedric Beaulac 
 
-LVM-MEGA
+LVM-MEGA : Test on gaussian mixture data
 """
 
 
@@ -69,9 +69,14 @@ pi=[0.3,0.3,0.4]
 mu=[[0,10],[-5,0],[5,2]]
 std=[[1,2],[0.5,0.5],[2,2]]
 
-sample = GMM_Gen2(n,pi,mu,std)
+data = GMM_Gen(n,pi,mu,std)
 
-plt.scatter(sample[:,0].numpy(),sample[:,1].numpy())
+plt.scatter(data[:,0].numpy(),data[:,1].numpy())
+
+ntest =500
+
+testdata = GMM_Gen(ntest,pi,mu,std)
+
 
 
 ###########################################
@@ -130,7 +135,8 @@ HDim = 10
 model =  PVAE(ObsDim,HDim,LDim).to(device)
 optimizer = optim.Adam(model.parameters())
 for epoch in range(1, 1000 + 1):
-    ptrain(args, model, device, sample, optimizer, epoch)
+    ptrain(args, model, device, data, optimizer, epoch)
+    ptest(args, model, device, testdata, epoch)
 
 
 #################################
@@ -138,10 +144,10 @@ for epoch in range(1, 1000 + 1):
 #################################
     
     
-nt = 5000
+ntz = 5000
 
 # Print new points
-NewPoint = np.random.normal(loc=np.zeros(LDim), scale=np.ones(LDim), size=(nt, LDim))
+NewPoint = np.random.normal(loc=np.zeros(LDim), scale=np.ones(LDim), size=(ntz, LDim))
 NewPoint = torch.tensor(NewPoint)
 NewPoint = NewPoint.type(torch.FloatTensor)
 
@@ -149,7 +155,7 @@ mux, logvarx = model.decode(NewPoint)
 
 varx = logvarx.exp_()
 varx = torch.diag_embed(varx[:,],0,1)
-param = torch.cat((mux.view(nt,1,2),varx),1)
+param = torch.cat((mux.view(ntz,1,2),varx),1)
 mvn = torch.distributions.multivariate_normal.MultivariateNormal(param[:,0,:],param[:,1:ObsDim+1,:])
 testsample = mvn.sample()
 
[…]

### 2021-01-30 main_MNIST.py

@@ -1,10 +1,10 @@
 # -*- coding: utf-8 -*-
 """
-Created on 22/01/21
+Created on 30/01/21
 
 @author: Cedric Beaulac 
 
-LVM-MEGA
+LVM-MEGA : Test on MNIST data 
 """
 
 
@@ -168,7 +168,10 @@ EzEx = np.mean(mux.detach().numpy(),0)
 
 EzEx/xbar
 
+# Inference model
+muz,logvarz = model.encoder(Data)
 
+ExEz = np.mean(muz.detach().numpy(),0)
 
 #################################
 # Second moment
@@ -204,8 +207,6 @@ RHS/LHS
 
 np.mean(RHS/LHS)
 
-
-
 #Inference model
 
 
@@ -217,8 +218,3 @@ for i in range(0,np.shape(npData)[0]):
     BigMatrix += V+E2
     
 SMZ = BigMatrix/(np.shape(npData)[0])
-
-
-muz,logvarz = model.encoder(sample)
-
-ExEz = np.mean(muz.detach().numpy(),0)

### 2021-01-30 utility.py

@@ -7,20 +7,15 @@ Clean place to drop function for LVS experiments
 """
 
 import torch
+import numpy as np
+
+
 ##################################  
-# GMM Generator
+# GMM_Gen: Gaussian Mixture Model Generator
+# inputs: n- number of observations, pi- probabilities of components
+# mu- mean of components and std- stand. dev. of component (no correlation atm)
 ##################################
-
 def GMM_Gen(n,pi,mu,std):
-    
-    ki = np.random.multinomial(1,pi,size=n) 
-    mui = np.matmul(ki,mu)
-    stdi = np.matmul(ki,std)
-    sample = np.random.normal(mui,stdi)
-            
-    return sample
-
-def GMM_Gen2(n,pi,mu,std):
     
     nk = torch.tensor(mu).size()[0]
     dim = torch.tensor(mu).size()[1]
@@ -35,4 +30,33 @@ def GMM_Gen2(n,pi,mu,std):
     sample = mvn.sample()
             
     return sample
+
+
+
+def power_iteration(A, num_simulations: int):
+    # Ideally choose a random vector
+    # To decrease the chance that our vector
+    # Is orthogonal to the eigenvector
+    b_k = np.random.rand(A.shape[1])
+
+    for _ in range(num_simulations):
+        # calculate the matrix-by-vector product Ab
+        b_k1 = np.dot(A, b_k)
+
+        # calculate the norm
+        b_k1_norm = np.linalg.norm(b_k1)
+
+        # re normalize the vector
+        b_k = b_k1 / b_k1_norm
+
+    return b_k
+
+
+def Frobenius(A):
+
+    A2 = np.dot(A,A) 
+    Trace = np.trace(A2)
+    Frob = sqrt(Trace)
+
+    return Frob
 

### 2021-02-04 Draft.tex
Lots of tex. modificaton

@@
 \usepackage{subcaption}
-\usepackage[margin=1.5in]{geometry}
+\usepackage[margin=1.25in]{geometry}
 \usepackage{float}
@@
 \usepackage[T1]{fontenc}
-\usepackage[sfdefault,scaled=.95,light]{FiraSans}
+\usepackage[sfdefault,scaled=1.05,light]{FiraSans}
 \usepackage{newtxsf}
@@
 \newcommand{\Ez}{\textbf{E}_z}
+\newcommand{\x}{\mathbf{x}}
+\newcommand{\z}{\mathbf{z}}
 
@@
 
-\title{Studying the gap between moment estimators to evaluate latent variable models}
+\title{Assessing latent variables models by studying the moment estimator gap}
 
@@
 
-\section{Latent variable models}
+When fitting supervised models, statisticians and computer scientist alike have come up with tons of metrics in order to evaluate the quality of their predictions and judge their adequacy, from simple mean-squared error to more general loss functions.
+However, in the context of unsupervised learning there is no direct measure of success and it can be difficult to assess the validity of the fitted model \cite{Hastie09}.
+
+\bigskip
+
+In this article we propose a new way to assess the quality of the fit a large family of unsupervised models.
+More precisely, we offer a diagnostic technique for parametric latent variable models such as Variational AutoEncoders (VAEs) \cite{Kingma13,Kingma17} and Gaussian Mixture Models (GMMs).
+Our proposed metrics evaluates the quality of the \textit{fitted model}; we compare the learned parameters of the unsupervised model with the observed data distribution the model is trying to capture.
+We do so by building two distinct moment estimators.
+
+\bigskip
+
+The technique we propose is fast to compute, works for a wide range of models and is built upon a rigorous mathematical formulation.
+It provides a new way to compare multiple models and behave accordingly to previously used heuristic techniques.
+
+\bigskip
+
+In the next section we establish the family of latent variable model suited for this metric.
+In section 3 we discuss the moments estimators we use before building the proposed evaluation metric in section 4.
+After we discuss related work in section 5 we present simulations and experiments in section 6.
+Finally, section 7 propose a wide range of future application of the ideas presented in this article such as the possibility of training models using the proposed metric.
+
+\section{Latent variable generative models}
+
+The diagnostic tool we propose is useful to assess the quality of the fit for a family of unsupervised latent variable models. In this section we define this family.
+
+\bigskip
+
+Assuming we have a data set $S= \{\mathbf{x}_1,...,\mathbf{x}_N\}$ consisting of $N$ observations of a $D$-dimensional variable $\mathbf{x}$.
+We assume $\x$ lives in the data-space $\mathcal{X}$ which is $D$-dimensional.
+We want to estimate the distribution of the random variable $\mathbf{x}$ but it is \textit{too complicated} to be captured by a simple distribution.
+Latent variable models suppose there exist a unobserved latent variable, say $\mathbf{z}$, that has a direct influence on the distribution of $\mathbf{x}$:
+\begin{align}
+p(\x) = \int_{z} p(\x|z)p(\z) dz
+\label{LVM}
+\end{align}
+where we assume $\z$ is in lives in a latent-space $\mathcal{Z}$ which is $M$-dimensional.
+The model proposed by equation \ref{LVM} is quite general but allows relatively complex marginal distributions over observed variables $\mathbf{x}$ to be expressed in terms of more tractable conditional distributions $p(\x|z)$ \cite{Bishop07}.
+Similarly, it leads to a tractable joint distribution as well:
+\begin{align}
+p(\x,\z) = p(\z)p(\x|z)
+\label{jointLVM}
+\end{align}
+
+and is quite often represented using a simple graph:
+\begin{figure}[h!]
+  \centering
+  \begin{tikzpicture}[->, semithick,scale = 0.9]
+  \tikzstyle{latent}=[fill=white,draw=black,text=black,style=circle,minimum size=1.0cm]
+  \tikzstyle{observed}=[fill=black!30,draw=black,text=black,style=circle,minimum size=1.0cm]
+
+  \node[latent] 		(A)	at (0,2)		{\large $\mathbf{z}$};
+
+  \node[observed]         (C)	at (0,0)		{\large $\mathbf{x}$};
+
+  \path (A) edge            node {} (C);
+
+\end{tikzpicture}
+  \caption{Graphical representation of latent variables models with joint distribution $p(\mathbf{x},\mathbf{z}) = p(\mathbf{z})p(\mathbf{x}|z)$.}
+  \label{genny}
+\end{figure}%
+These models are \textit{generative models} because learning $p(\x,\z)= p(\z)p(\x|z)$ allows us to generated new samples of $\x$ using ancestral sampling.
+What makes this model probabilistic is that the mapping from $\z$
+ to $\x$ is not a deterministic function $f: \mathcal{Z} \rightarrow \mathcal{X}$ but instead a \textit{probabilistic mapping} from $\mathcal{Z}$ to the parameter space $\Theta$, where $\Theta$ is the parameter space of $p_\theta(\x|z)$; $\theta \in \Theta$.
+
+\bigskip
+
+When training or fitting such models we are training the function $f:\mathcal{Z} \rightarrow \Theta$ in ways to maximize the likelihood of the data set $S$ under the model of equation \ref{LVM}.
+This mapping $f$ explains the effect of $\z$ on $\x$ ans is at the centre of latent variable models.
+Learning this function $f$ is the main challenge of optimizing latent variable model and the Expectation-Maximization algorithm (EM) or variational inference are common strategies to fit learn such function.
+
+\bigskip
+
+Usually $p_\theta(\x|z)$ is a simple parametric distribution and the latent variable is in charge of increasing the complexity of $p_\theta(\x)$.
+Additionally, the function $f$ can take many forms, from simple linear combination to neural network functions.
+We will use $f(\z)$ interchangeably with $f$ but also sometimes the \textit{distribution parameters} it outputs directly.
+For instance, assuming $p_\theta(\x|z) = \text{Poisson}(\lambda)$ then $f: \mathcal{Z} \rightarrow \mathbb{R}^+$ and we might use $f(\z)$ and $\lambda(\z)$ interchangeably.
+
+\subsection{Gaussian Mixture Models}
+
+The GMM is a member of the family latent variable generative models we just described.
+For a GMM with $K$-components we define $\z$ as a k-class categorical variable, $\z \in \{1,...,K\} = \mathcal{Z}$, $p(\z)$ is a categorical distribution where we define $\pi_j$ as $p(\z = j)$ with $\sum_{j=1}^K \pi_j =1$.
+Finally, setting $p(\x|z=j) = N(\mu_j,\sigma_j)$ leads to a GMM:
+\begin{align}
+p(\x) = \sum_{j=1}^K \pi_j p(\x|z=j) 
+\label{LVM}
+\end{align}
+In this situation, $f$ map the latent variable to a pair of \textit{distribution parameters}, $\mu$ and $\sigma$, $f: \{1,...,K\} \rightarrow \mathbb{R}\times\mathbb{R}^+$. Similarly
+
+\subsection{Probabilistic Principal Component Analysis}
+
+\subsection{Variational AutoEncoders}
 
@@
 
-The idea is to build two different estimators of the same value. One uses observed data and the other uses the proposed generative model.
+We propose to build two different estimators of the same value.
+One uses observed data and the other uses the proposed generative model.
+What makes our model estimator different is that we do not generate points from the generative model but instead we rely on a simple probability identity to build the \textit{model} estimator.
+The estimator
 
@@
 
-The Law of Total Variance :
+To do, let us introduce the well-know Law of Total Variance :
 
@@
 
-Notice the second term is equal to :
+and notice the second term is equal to :
 
@@
 We call the LHS the data estimate (DE).
-The right and side of equation \ref{rhslhs} can be estimated using every components of the proposed generative model; we estimate it via a Monte Carlo sample of the prior $p_\theta(z)$ and both $\Vx(X|Z)$ and $\Ex(X|Z)$ which completely defined $p_\theta(x|z)$ contribute to this term:
+The right and side of equation \ref{rhslhs} can be estimated using every components of the proposed generative model; we estimate it via a Monte Carlo sample of the prior $p_\theta(z)$ and both $\Vx(X|Z)$ and $\Ex(X|Z)$ which completely defined $p_\theta(x|z)$ contribute to this term.
 
@@
 
-where $z_i \sim p_\theta(z)$, MLP$_1$ is the multilayer perceptron that return $\Ex(X|Z)$ and MLP$_2$ is the multilayer perceptron that return $\Vx(X|Z)$. This is the forward model estimates (FME).
+where $z_i \sim p_\theta(z)$, MLP$_1$ is the multilayer perceptron that return $\Ex(X|Z)$ and MLP$_2$ is the multilayer perceptron that return $\Vx(X|Z)$.
+This is the forward model estimates (FME).
+Notice that this estimator does not require we sample from $p_\theta(x|z)$ and uses directly the \textit{parameters} of this distribution instead.
+It is very fast to sample a large amount of $z's$ and additionally, because this is a rather simple Monte Carlo sample, we can compute its estimated standard error ...
 
@@
 
-The estimates from equation \ref{lhs} should be similar to those from equation \ref{rhs}.
-Right now, we take the element-wise ratio of both these estimates and if the propose generative models does capture well the second moment of $X$ we expect elements of this matrix to be close to one.
-
-\bigskip
-
-Based on simulation we observe that the generative model tends to underestimate the second moment.
+The estimates from equation \ref{lhs} should be similar to those from equation \ref{rhs}. We propose to compute the matrix norm of the difference between those two estimators.
 
@@
 
-Know that we can do something similar for the first moment :
+We can do something similar for the first moment :
 
@@
 where we can estimate the left-hand side with $\bar{x}$ and the right-hand side with $\frac{1}{m} \sum_{i=1}^m \text{MLP}_1(z_i)$ where $z_i \sim p_\theta(z)$.
-However, based on our previous experiment we do not believe capturing the first moment is an issue vor VAE at the current moment and the RHS estimate only account for a subset of the estimated parameters (it does not include the parameters of MLP$_2$).
+However, based on our previous experiment we do not believe capturing the first moment is an issue for most latent variable model.
+Additionally, the RHS estimate only account for a subset of the estimated parameters (it does not include the parameters of MLP$_2$) which makes the gap between those two estimators less interesting to study.
+However, to get a complete picture of the models proficient at estimating the observed-data distribution we include a first moment analysis as well.
 
-\bigskip
-
-The first objective is to use such strategies as diagnostic tools to evaluate the fit of latent variable model. A follow-up project is to use those as part of the optimization procedure.
-
-\section{Gap between moment estimators}
+\section{MEGA (Moment Estimators GAp)}
 
 For latent variable with a define forward mechanism, such as pPCA, GMM, HMM and VAE we can assess the ability of the forward model to capture the moment of the observation by comparing the DE with the FME.
-From now we refer to the object we are studying as MEGA (Moment Estimators GAp).
+From now we refer to the object we study, the matrix that is the difference between our two second moment estimator, as MEGA (Moment Estimators GAp) which we identify as $M$.
 
@@
 We want to use a norm that looks at the global properties of the matrix and fortunately Rigollet \cite{rigollet15} introduces multiple norms that do so.
+To begin we can use norms inspired by vector norms, givent the vector $v$ if $|v|_q$ is the following vector norm:
+
+\begin{align}
+|v|_q = (\sum_i |v_i|^q)^{(1/q)},
+\label{vecnorm}
+\end{align}
+
+then its matrix equivalent is $|M|_q$:
+
+\begin{align}
+|M|_q = (\sum_{ij} |M_{ij}|^q)^{(1/q)},
+\label{matnorm}
+\end{align}
+
+$q=2$ is a special case call the Frobenius norm:
+
+\begin{align}
+|M|_2 = |M|_F = (\sum_{ij} |M_{ij}|^2)^{(1/2)} = \sqrt{\text{Tr}(M^TM)}.
+\label{Frobe}
+\end{align}
+
+where Tr is the trace operator that sums the elements of the diagonal of the input matrix.
+This norm is also a member of the Schatten $q$-norms (for $q=2$) which is a family of matrix norms defined as the vector norm of equation \ref{vecnorm} for the singular values of the matrix.
+Since we work with second moment estimators, our matrix $M$ is a squared matrix (dimension $m \times m$) and positive semidefinite and thus the singular values of $M$ are equal to its eigenvalues.
+We identify the vector of eigenvalues as $\lambda$.
+Thus the  Schatten $q$-norm for matrix $M$ is $||M||_q=|\lambda|_q$.
+
+\bigskip
+
+Another member of this family is consider, when $q=\infty$ we define $||M||_\infty = \lambda_{max} = ||M||_{op}$ and this was named the operator norm.
+
+\bigskip
+
+In random matrix theory and in matrix estimation these two norms appear frequently and consequently are well-know in the statistical research community.
+For instance, in covariance matrix estimation it is possible, under mild assumptions, to bound the operator norm of the difference between the true covariance matrix and simple estimators \cite{rigollet15}.
+Because of the popularity and the known properties of both the Frobenius norm and the operator norm, they are legitimate options to measure the MEGA.
+
+\bigskip
+
+Additional, both norms can be computed quite efficiently.
+To compute to Frobenius norm, we need square of the MEGA , $m^2$ operations, and then sum its diagonal which results in $m^2+m$ operations.
+The largest eigenvalue of a matrix can be estimated with the von Mises iteration \cite{mises29}, where each of the $p$ iterations will require $m^2$ operations resulting in $pm^2$ operations in total.
 
@@
 
+\section{Related Works}
+
 \section{Experiments and results}
 
-\section{Training with moment estimator matching}
+\section{Future work}
+\subsection{Extension to a wider class of probabilistic graphical models}
+\subsection{Training by minimizing MEGA}
 
 \section{Conclusion}
+
+\section*{Ackowledgement}
+
+Yanbo Tang, Michael Lalancette, Jeffrey S. Rosenthal and Renaud Alie.
 

### 2021-02-04 main_GMM.py

@@ -40,6 +40,7 @@ from sklearn.ensemble import VotingClassifier
 from sklearn.metrics import accuracy_score
 from sklearn.preprocessing import LabelEncoder
 from sklearn.preprocessing import OneHotEncoder
+from sklearn.mixture import GaussianMixture
 
 import os
 
@@ -56,6 +57,7 @@ import matplotlib.pyplot as plt
 
 from utility import *
 from NNutility import *
+from pyppca import *
 
 
 
@@ -125,28 +127,35 @@ kwargs = {'num_workers': 1, 'pin_memory': True} if use_cuda else {}
 
 
 ###########################################
-# Defining model and training
+# Defining models and training
 ###########################################
 
-
-LDim = 2
+#VAE
+LDim = 5
 HDim = 10
 
 model =  PVAE(ObsDim,HDim,LDim).to(device)
 optimizer = optim.Adam(model.parameters())
-for epoch in range(1, 1000 + 1):
+for epoch in range(1, 500 + 1):
     ptrain(args, model, device, data, optimizer, epoch)
     ptest(args, model, device, testdata, epoch)
+    
+    
+#GMM
+gm = GaussianMixture(n_components=3, random_state=0).fit(data)
+
+    
+
 
 
 #################################
 # Testing process (Garbage)
 #################################
     
-    
+ 
 ntz = 5000
 
-# Print new points
+#VAE 
 NewPoint = np.random.normal(loc=np.zeros(LDim), scale=np.ones(LDim), size=(ntz, LDim))
 NewPoint = torch.tensor(NewPoint)
 NewPoint = NewPoint.type(torch.FloatTensor)
@@ -157,10 +166,20 @@ varx = logvarx.exp_()
 varx = torch.diag_embed(varx[:,],0,1)
 param = torch.cat((mux.view(ntz,1,2),varx),1)
 mvn = torch.distributions.multivariate_normal.MultivariateNormal(param[:,0,:],param[:,1:ObsDim+1,:])
-testsample = mvn.sample()
+vaetestsample = mvn.sample()
 
-#plt.scatter(mux[:,0].detach().numpy(),mux[:,1].detach().numpy())
-plt.scatter(testsample[:,0].numpy(),testsample[:,1].numpy())
[…]

### 2021-02-04 utility.py

@@ -8,7 +8,9 @@ Clean place to drop function for LVS experiments
 
 import torch
 import numpy as np
+from numpy import *
 
+import pandas as pd
 
 ##################################  
 # GMM_Gen: Gaussian Mixture Model Generator
@@ -32,7 +34,10 @@ def GMM_Gen(n,pi,mu,std):
     return sample
 
 
-
+##################################  
+# power_iteration (Von Mises iteration)
+# copied from https://en.wikipedia.org/wiki/Power_iteration
+##################################
 def power_iteration(A, num_simulations: int):
     # Ideally choose a random vector
     # To decrease the chance that our vector
@@ -52,7 +57,10 @@ def power_iteration(A, num_simulations: int):
     return b_k
 
 
-def Frobenius(A):
+##################################  
+# frobnorm : Frobenius norm of matrix A
+##################################
+def frobnorm(A):
 
     A2 = np.dot(A,A) 
     Trace = np.trace(A2)
@@ -60,3 +68,13 @@ def Frobenius(A):
 
     return Frob
 
+
+
+def ppca(x,d,dia):
+    
+    #Initialize parameters
+    
+    n,D = shape(x)
+    W = np.random.normal(0,1,[D,d])
+
+    return 

### 2021-02-05 utility.py
ppca (incomplete)

@@ -12,6 +12,11 @@ from numpy import *
 
 import pandas as pd
 
+import torch
+import torch.nn.init as init
+import torch.nn.functional as F
+import torch.nn as nn
+
 ##################################  
 # GMM_Gen: Gaussian Mixture Model Generator
 # inputs: n- number of observations, pi- probabilities of components
@@ -69,12 +74,39 @@ def frobnorm(A):
     return Frob
 
 
+##################################  
+# ppca: Training a probabilistic PCA with EM (scales better)
+##################################
+def ppca(x,m,dia,ite):
+    
 
-def ppca(x,d,dia):
+    #n,D are fixed by the problem
+    #mu being the mean is the ML estimator, we fix it first 
+    n,d = shape(x)
+    mu = np.mean(x,0)
+    xd = torch.tensor(x-mu)
+
     
-    #Initialize parameters
+    #Random Initilization of parameters in need of training
+    W = torch.tensor(np.random.normal(0,1,[d,m]))
+    sig = torch.tensor(exp(np.random.normal(0,0.1,[1])))
     
-    n,D = shape(x)
-    W = np.random.normal(0,1,[D,d])
+    #Initialize matrix M
+    M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
+
+    
+    for i in range(0,ite):
+        
+        #Setting stuff up
+        Minv = torch.inverse(M)
+
+        
+        #E Step
+        Ez =  torch.matmul(torch.matmul(Minv,torch.transpose(W,0,1)),torch.transpose(xd,0,1))
+        Ez = Ez.view(100,2,1)
+        Ezt = Ez.view(100,1,2)
+        Ezz = sig*Minv + torch.matmul(Ez,Ezt)
+        
+        #M Step
 
     return 

### 2021-02-08 Draft.tex
ppca (incomplete)
Completed ppca, tiny bits of article update

@@
 
-\section{Intro}
+\section{Introduction} \label{intr}
 
@@
 In the next section we establish the family of latent variable model suited for this metric.
-In section 3 we discuss the moments estimators we use before building the proposed evaluation metric in section 4.
-After we discuss related work in section 5 we present simulations and experiments in section 6.
-Finally, section 7 propose a wide range of future application of the ideas presented in this article such as the possibility of training models using the proposed metric.
+After we discuss related work in section \ref{rewo}, we discuss the moments estimators used in section \ref{moes}.
+In section \ref{mega}, we present our metric.
+We show how this metric perform with simulations in section \ref{expe}.
+Finally, section \ref{fuwo} propose a wide range of future application of the ideas presented in this article such as the possibility of training models using the proposed metric.
 
-\section{Latent variable generative models}
+\section{Latent variable generative models}\label{lvgm}
 
-The diagnostic tool we propose is useful to assess the quality of the fit for a family of unsupervised latent variable models. In this section we define this family.
+The diagnostic tool we propose is useful to assess the quality of the fit for a family of unsupervised latent variable models.
+In this section we define this family, the Latent variable generative models (LVGMs).
 
@@
  to $\x$ is not a deterministic function $f: \mathcal{Z} \rightarrow \mathcal{X}$ but instead a \textit{probabilistic mapping} from $\mathcal{Z}$ to the parameter space $\Theta$, where $\Theta$ is the parameter space of $p_\theta(\x|z)$; $\theta \in \Theta$.
+We will call $p_\theta(\x|z)$ the emission distribution or observation distribution interchangeably.
 
@@
 Learning this function $f$ is the main challenge of optimizing latent variable model and the Expectation-Maximization algorithm (EM) or variational inference are common strategies to fit learn such function.
+In most cases, $p(\z)$ is assumed to be known and fixed but in some cases the parameters of $p(\z)$ are estimated as well.
 
@@
 We will use $f(\z)$ interchangeably with $f$ but also sometimes the \textit{distribution parameters} it outputs directly.
-For instance, assuming $p_\theta(\x|z) = \text{Poisson}(\lambda)$ then $f: \mathcal{Z} \rightarrow \mathbb{R}^+$ and we might use $f(\z)$ and $\lambda(\z)$ interchangeably.
+For the majority of emission distribution we consider, there exist a simple mapping from the parameters of the distribution to its expectation and its variance.
+
+\bigskip
+
+For instance, assuming the emission distribution is Poisson: $p_\theta(\x|z) = \text{Poisson}(\lambda)$ then $f: \mathcal{Z} \rightarrow \mathbb{R}^+$ and we might use $f(\z)$ and $\lambda(\z)$ interchangeably.
+Additionally, $\Ex[\x|\z] = \lambda(\z)$ and $ \Vx[\x|\z] = \lambda(\z)$.
+
+\subsection{Probabilistic Principal Component Analysis}
+
+The Probabilistic Principal Component Analysis (pPCA) \cite{tipping99,Bishop07} is a member of the LVGM family we just described where $p(\z)$ is assumed to be a normal distribution $N(0,I)$.
+We also assume the emission distribution to be Normal: $p(\x|\z) = N(W\z+\mu,\sigma^2I)$.
+In this formulation, we see that $f: \mathbb{R}_M \rightarrow \mathbb{R}_D$ is a linear function that maps the latent variable $\z$ to $\Ex[\x|\z]$.
+Outside of estimating $W$ and $\mu$ as part of $f$ the model also estimate $\sigma^2$ though it is not a function of $z$.
+However it is function of $D$ and $M$, the dimension of $\mathcal{X}$ and $\mathcal{Z}$.
+
+\bigskip
+
+The parameters of pPCA can be fit analytically as the solution of a direct maximization of the likelihood or with the EM algorithm.
+
+\subsection{Variational AutoEncoders}
 
@@
 
-The GMM is a member of the family latent variable generative models we just described.
 For a GMM with $K$-components we define $\z$ as a k-class categorical variable, $\z \in \{1,...,K\} = \mathcal{Z}$, $p(\z)$ is a categorical distribution where we define $\pi_j$ as $p(\z = j)$ with $\sum_{j=1}^K \pi_j =1$.
@@
 \end{align}
-In this situation, $f$ map the latent variable to a pair of \textit{distribution parameters}, $\mu$ and $\sigma$, $f: \{1,...,K\} \rightarrow \mathbb{R}\times\mathbb{R}^+$. Similarly
+In this situation, $f$ map the latent variable to a pair of \textit{distribution parameters}, $\mu$ and $\sigma$, $f: \{1,...,K\} \rightarrow \mathbb{R}\times\mathbb{R}^+$.
+In this particular cases  $\Ex[\x|\z] = f_1(\z)$ where $f_1$ is the \textit{first} output of $f(\z)$ and $\Vx[\x|\z] = f_2(\z)$.
 
-\subsection{Probabilistic Principal Component Analysis}
+\bigskip
 
-\subsection{Variational AutoEncoders}
+The GMM is a special case of LVGM where we also estimate the parameters $\{\pi_j : j \in 1,...K\}$ of $p_(\z)$. A GMM is usually train using the EM algorithm.
 
-\section{Moment estimators}
+\section{Related Works} \label{rewo}
+
+We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature there is no good metric for such thing at the moment.
+
+\bigskip
+
+In the literature the likelihood of the data $S$ under the fitted model is used CITE, however,it has been shown CITE that sometimes a model can reach very high-likelihood even for terrible fit.
+For instance, Bishop et al.
+\cite{Bishop07} demonstrate that for
+
+\section{Moment estimators} \label{moes}
 
@@
 \begin{align}
-\Vx(X) = \Ez[\Vx(X|Z)] + \Vz[\Ex(X|Z)]
+\Vx(\x) = \Ez[\Vx(\x|\z)] + \Vz[\Ex(\x|\z)]
 \end{align}
@@
 \Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
-&\approx \frac{1}{m} \sum_{i=1}^m ( \text{diag}(\text{MLP}_2(z_i)) + \text{MLP}_1(z_i)^T\text{MLP}_1(z_i))
+&\approx \frac{1}{m} \sum_{i=1}^m \left[\Vx(\x|\z=z_i) + \Ex(\x|\z=z_i)^T\Ex(\x|\z=z_i)\right]
 \label{rhs}
@@
 
-where $z_i \sim p_\theta(z)$, MLP$_1$ is the multilayer perceptron that return $\Ex(X|Z)$ and MLP$_2$ is the multilayer perceptron that return $\Vx(X|Z)$.
+where $z_i \sim p_(\z)$, and $\Ex(\x|\z=z_i)$ and  $\Vx(\x|\z=z_i)$ are expressed as functions of $f(z_i)$.
+Thus this estimator relies on all the components of the fitted LVGM: $p(\z)$ and $f(\z)$.
 This is the forward model estimates (FME).
-Notice that this estimator does not require we sample from $p_\theta(x|z)$ and uses directly the \textit{parameters} of this distribution instead.
+Notice that this estimator does not require we sample from $p_\theta(x|z)$ and uses directly the estimated parameters of the emission distribution.
 It is very fast to sample a large amount of $z's$ and additionally, because this is a rather simple Monte Carlo sample, we can compute its estimated standard error ...
@@
 \Ex[X] = \Ez[\Ex(X|Z)] 
-\label{rhs}
+\label{1stm}
 \end{align}
@@
 
-\section{MEGA (Moment Estimators GAp)}
+\section{MEGA: Moment Estimators GAp} \label{mega}
 
@@
 
-\section{Related Works}
+\section{Experiments and results} \label{expe}
 
-\section{Experiments and results}
-
-\section{Future work}
+\section{Future work} \label{fuwo}
 \subsection{Extension to a wider class of probabilistic graphical models}
@@
 
-\section{Conclusion}
+\section{Conclusion} \label{conc}
 

### 2021-02-08 utility.py

@@ -92,12 +92,13 @@ def ppca(x,m,dia,ite):
     sig = torch.tensor(exp(np.random.normal(0,0.1,[1])))
     
     #Initialize matrix M
-    M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
+    #M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
 
-    
+    # ite iterations of the EM algorithm
     for i in range(0,ite):
         
         #Setting stuff up
+        M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
         Minv = torch.inverse(M)
 
         
@@ -108,5 +109,12 @@ def ppca(x,m,dia,ite):
         Ezz = sig*Minv + torch.matmul(Ez,Ezt)
         
         #M Step
-
-    return 
+        W1 = torch.sum(torch.matmul(xd.view(100,5,1),Ezt),0)
+        W2 = torch.sum(Ezz,0)
+        W = torch.matmul(W1,W2)
+        s1 = torch.sum(xd.pow(2),1).pow(0.5)
+        s2 = torch.matmul(xd.view(100,1,5),torch.matmul(W.repeat(100,1,1),Ez)).view(100)
+        s3 = torch.diagonal(torch.matmul(Ezz,torch.matmul(torch.transpose(W,0,1),W).repeat(100,1,1)), dim1=-2, dim2=-1).sum(-1)
+        sig = torch.sum(s1+s2+s3)/(n+d)
+        
+    return mu,W,sig

### 2021-08-28 Draft.tex
Pushing both MEGA files

@@
 
-\title{Assessing latent variables models by studying the moment estimator gap}
+\title{MEGA: New metrics to evaluate latent variable generative models}
 
@@
 \maketitle
+\begin{abstract}
+In this article we introduce a new metric built to measure how latent variable models, such as Gaussian Mixture Models or Variational AutoEncoders, capture the first and second moments of data sets.
+We propose to compare the data moments to a moment estimator that uses both components of latent variable models.
+In order to make the gap between those two estimator digestible, we propose using the Frobenius norm of this gap to measure if the latent variable model did capture the moments of the data set.
+We discuss our metric behaviour under various scenarios and demonstrate it on multiple data sets with various latent variable models.
+Finally, we discuss various extensions: we discuss how to extend this metric to sequential latent variable model such as Hidden Markov Models and State-Space models, how to extend this metric to higher moments and the possibility of training model by minimizing this metric.
+\bigskip
+
+%\noindent
+
+\textbf{Keywords} : Moment estimators, Latent Variable models, Frobenius norm, Gaussian Mixture Models, Variational Auto-Encoders
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
+\end{abstract}
+
+\pagebreak
 
@@
 
+The VAE is also a member of the LVGM family.
+It is assumed that $\z$ is a continuous variable where $p(\z)$ is assumed to be $N(0,I)$.
+$p(\x|\z)$ can be any parametric distribution where $f(\z)$ outputs the parameters of this distribution.
+For instance, if $p(\x|\z)$ is normal then $f(\z)$ will output a mean and a variance parameter, $f: \mathcal{R}_m \rightarrow \mathcal{R}_d \times \mathcal{R}_(d,d)^+$.
+
+\bigskip
+
+One novelty of VAEs is that the function $f$ proposed is much more flexible than a linear combination; it is a neural network.
+In turns, this makes the posterior distribution $p(\z|\x)$ intractable and which prevent the model from being fitted by the EM algorithm.
+The solution proposed is to assume a variational distribution $q(\z|\x)$ and optimize the system by maximizing the Evidence Lower BOund (ELBO), a lower bound of the observed-data log-likelihood.
+
 \subsection{Gaussian Mixture Models}
@@
 
-We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature there is no good metric for such thing at the moment.
+We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature the lack of a reliable metric to asses latent variable models is obvious.
+At the moment, researchers rely either on the likelihood of the data under the fitted model or the heuristic analysis of generated data point.
+Both techniques have their fair share of problems.
+
+\bigskip
+
+\cite{Kingma13,
 
@@
 \begin{align}
-\Ex[X^2] = \Vx(X)+ (\Ex[X])^2 \approx \frac{\sum_{i=1}^n(X_i-\bar{X})^T(X_i-\bar{X})}{n-1} + \bar{X}^T\bar{X} = \frac{\sum_{i=1}^n X_i^TX_i}{n-1}
+\text{DE} = \Ex[X^2] = \Vx(X)+ (\Ex[X])^2 \approx \frac{\sum_{i=1}^n(X_i-\bar{X})^T(X_i-\bar{X})}{n-1} + \bar{X}^T\bar{X} = \frac{\sum_{i=1}^n X_i^TX_i}{n-1}
 \label{lhs}
@@
 \begin{align}
-\Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
+\text{FME}= \Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
 &\approx \frac{1}{m} \sum_{i=1}^m \left[\Vx(\x|\z=z_i) + \Ex(\x|\z=z_i)^T\Ex(\x|\z=z_i)\right]
@@
 However, to get a complete picture of the models proficient at estimating the observed-data distribution we include a first moment analysis as well.
+
+\subsection{Additional justification}
+
+\subsection{More moments}
 
@@
 
-Additional, both norms can be computed quite efficiently.
-To compute to Frobenius norm, we need square of the MEGA , $m^2$ operations, and then sum its diagonal which results in $m^2+m$ operations.
+Finally, the metrics we propose to evaluate the quality of the fit for the second moment is:
+
+\begin{align}
+\text{MEGA-F} &= |\text{DE-FME}|_F. \\
+\text{MEGA-Op} &= |\text{DE-FME}|_{op}. \\
+\label{MEGAs}
+\end{align}
+
+\subsection{Implementation of the selected norm}
+
+Implementing the Frobenius norm is quite straight forward. To compute to Frobenius norm, we need square of the MEGA , $m^2$ operations, and then sum its diagonal which results in $m^2+m$ operations.
+
+\bigskip
+
 The largest eigenvalue of a matrix can be estimated with the von Mises iteration \cite{mises29}, where each of the $p$ iterations will require $m^2$ operations resulting in $pm^2$ operations in total.
 
-\subsection{Implementation of the selected norms}
-
-\section{Experiments and results} \label{expe}
+\section{Experiments: Assessing VAE's ability to capture second order properties} \label{expe}
 

### 2021-08-28 MEGA_Numpy.py

@@ -1,0 +1,65 @@
+# -*- coding: utf-8 -*-
+"""
+Created on 12/02/21
+
+@author: Cedric Beaulac 
+
+LVM-MEGA : Implementation for MEGA1 and MEGA2
+"""
+import numpy as np
+##################################  
+# frobnorm : Frobenius norm of the matrix A
+##################################
+def frobnorm(A):
+
+    A2 = np.dot(A,A) 
+    Trace = np.trace(A2)
+    Frob = np.sqrt(Trace)
+
+    return Frob
+
+##################################  
+# frobvect : Frobenius norm of the vector a
+##################################
+def frobvect(a):
+
+    Frob = np.sqrt(np.sum(np.power(a,2)))
+
+    return Frob
+
+
+
+#################################
+# MEGA: Returns MEGA1 and MEGA2 (using Frob norm)    
+# Inputs: data (Data set), mean (sample of E[x|z]), var (sample of V[x|z])
+# Output : MEGA1 (MEGA for first moment) and MEGA2 (MEGA for second moment)
+#################################
+def MEGA(data,mean,var):
+    
+    #Initialize MEGA
+    MEGA = np.zeros(2)
+    nz = np.shape(mean)[0]
+    d = np.shape(mean)[1]
+
+
+    
+    #First moment MEGA
+    #DE
+    xbar = np.mean(data,0)
+    #FME
+    EzEx = np.mean(mean,0)
+    MEGA[0] = frobvect(EzEx-xbar)
+
+    # Second moment MEGA
+    # LHS (DE)
+    S2 = np.cov(np.transpose(data))
+    xbar2 = np.outer(xbar,np.transpose(xbar))
+    LHS = S2+xbar2
+    # RHS (FME)
+    E2 = np.matmul(mean.reshape(-1,np.shape(mean)[1],1),mean.reshape(-1,1,np.shape(mean)[1]))
+    RHS = np.mean(E2+var,0)
+    
+    MEGA[1] = frobnorm(LHS-RHS)
+
+    
+    return MEGA

### 2021-08-28 MEGA_Pytorch.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+"""
+Created on 27/03/21
+
+@author: Cedric Beaulac 
+
+LVM-MEGA : Implementation for MEGA1 and MEGA2 in PyTorch
+"""
+import torch
+import numpy as np
+##################################  
+# frobnorm : Frobenius norm of the matrix A
+##################################
+def frobnorm(A):
+
+    A2 = torch.matmul(A,A) 
+    Trace = torch.trace(A2)
+    Frob = torch.sqrt(Trace)
+
+    return Frob
+
+##################################  
+# frobvect : Frobenius norm of the vector a
+##################################
+def frobvect(a):
+
+    Frob = torch.sqrt(torch.sum(torch.pow(a,2)))
+
+    return Frob
+
+
+
+#################################
+# MEGA: Returns MEGA1 and MEGA2 (using Frob norm)    
+# Inputs: data (Data set,ndarray), mean (sample of E[x|z],tensor), var (sample of V[x|z],tensor)
+# Output : MEGA1 (MEGA for first moment) and MEGA2 (MEGA for second moment)
+#################################
+def MEGA(data,mean,var):
+    
+    #Initialize MEGA
+    MEGA = torch.zeros(2)
+    nz = mean.size()[0]
+    d = mean.size()[1]
+
+
+    
+    #First moment MEGA
+    #DE
+    xbar = torch.mean(data,0)
+    #FME
+    EzEx = torch.mean(mean,0)
+    MEGA[0] = frobvect(EzEx-xbar)
+
+    # Second moment MEGA
+    # LHS (DE)
+    S2 = np.cov(np.transpose(data))
+    xbar2 = np.outer(xbar,np.transpose(xbar))
+    LHS = torch.tensor(S2+xbar2)
+    # RHS (FME)
+    E2 = torch.matmul(mean.reshape(-1,mean.size()[1],1),mean.reshape(-1,1,mean.size()[1]))
+    RHS = torch.mean(E2+var,0)
+    
+    MEGA[1] = torch.sqrt(frobnorm(LHS-RHS))
+
+    
+    return MEGA

### 2021-08-28 main_GMM.py

@@ -57,7 +57,7 @@ import matplotlib.pyplot as plt
 
 from utility import *
 from NNutility import *
-from pyppca import *
+
 
 
 
@@ -66,10 +66,10 @@ from pyppca import *
 ####################################
 
 
-n=5000
+n=1000
 pi=[0.3,0.3,0.4]
-mu=[[0,10],[-5,0],[5,2]]
-std=[[1,2],[0.5,0.5],[2,2]]
+mu=[[0,-5],[-2,0],[4,2]]
+std=[[1,2],[2,2],[2,2]]
 
 data = GMM_Gen(n,pi,mu,std)
 
@@ -78,6 +78,27 @@ plt.scatter(data[:,0].numpy(),data[:,1].numpy())
 ntest =500
 
 testdata = GMM_Gen(ntest,pi,mu,std)
+
+
+
+####################################
+# Generate data sets
+####################################
+
+
+n=1000
+pi=[0.3,0.3,0.4]
+mu=[[0,-5,0,-10],[-2,0,5,5],[4,2,0,-10]]
+std=[[1,2,1,1],[2,2,0.5,10],[2,2,2,2]]
+
+data = GMM_Gen(n,pi,mu,std)
+
+plt.scatter(data[:,3].numpy(),data[:,1].numpy())
+
+ntest =500
+
+testdata = GMM_Gen(ntest,pi,mu,std)
+
 
 
 
@@ -136,7 +157,7 @@ HDim = 10
 
 model =  PVAE(ObsDim,HDim,LDim).to(device)
 optimizer = optim.Adam(model.parameters())
-for epoch in range(1, 500 + 1):
+for epoch in range(1, 1000 + 1):
     ptrain(args, model, device, data, optimizer, epoch)
     ptest(args, model, device, testdata, epoch)
     
@@ -144,16 +165,19 @@ for epoch in range(1, 500 + 1):
 #GMM
 gm = GaussianMixture(n_components=3, random_state=0).fit(data)
 
-    
+
+#pPCA
+ldim =2 
+mu,W,sig2 = ppca(data,ldim,5000)
 
 
-
+mu,W,sig2 = ppca_ll(data,ldim)
 #################################
-# Testing process (Garbage)
+# Generate points from fitted models
 #################################
     
  
-ntz = 5000
+ntz = 1000
 
 #VAE 
[…]

### 2021-08-28 main_MNIST.py

@@ -7,7 +7,6 @@ Created on 30/01/21
 LVM-MEGA : Test on MNIST data 
 """
 
-
 ####################################
 # Import the good stuff (bunch of libraries)
 ####################################
@@ -40,6 +39,7 @@ from sklearn.ensemble import VotingClassifier
 from sklearn.metrics import accuracy_score
 from sklearn.preprocessing import LabelEncoder
 from sklearn.preprocessing import OneHotEncoder
+from sklearn.mixture import GaussianMixture
 
 import os
 
@@ -59,27 +59,34 @@ from NNutility import *
 
 
 
+
 ####################################
 # Generate data sets
 ####################################
 
 
-n=5000
-pi=[0.3,0.3,0.4]
-mu=[[0,10],[-5,0],[5,2]]
-std=[[1,2],[0.5,0.5],[2,2]]
+big_loader = torch.utils.data.DataLoader(
+    datasets.MNIST('../data', train=True, download=True,
+                   transform=transforms.ToTensor()),batch_size=20000)
+for batch_idx, (x,y) in enumerate(big_loader):
+    x,y =x,y
 
-sample = GMM_Gen2(n,pi,mu,std)
+test_loader = torch.utils.data.DataLoader(
+    datasets.MNIST('../data', train=False, transform=transforms.ToTensor()),
+    batch_size=500, shuffle=True, **kwargs)
+for batch_idx, (tx,ty) in enumerate(big_loader):
+    tx,ty =tx,ty
 
-plt.scatter(sample[:,0].numpy(),sample[:,1].numpy())
+data = x.view(-1,784)
 
+testdata = tx.view(-1,784)
 
 ###########################################
 # Fixing arguments
 ###########################################
 
 
-ObsDim = shape(mu)[1]
+ObsDim = shape(data)[1]
 BSize = 500
 NB = int(n/BSize)
 
@@ -120,28 +127,39 @@ kwargs = {'num_workers': 1, 'pin_memory': True} if use_cuda else {}
 
 
 ###########################################
-# Defining model and training
+# Defining models and training
 ###########################################
 
-
-LDim = 2
-HDim = 10
+#VAE
+LDim = 5
[…]

### 2021-08-28 utility.py

@@ -74,47 +74,84 @@ def frobnorm(A):
     return Frob
 
 
+
+def frobvect(a):
+
+
+    frob = sqrt(np.sum(np.power(a,2)))
+
+    return Frob
+
+
+
 ##################################  
 # ppca: Training a probabilistic PCA with EM (scales better)
 ##################################
-def ppca(x,m,dia,ite):
+def ppca(x,m,ite):
     
 
     #n,D are fixed by the problem
     #mu being the mean is the ML estimator, we fix it first 
     n,d = shape(x)
-    mu = np.mean(x,0)
-    xd = torch.tensor(x-mu)
+    mu = torch.mean(x,0).float()
+    xd = torch.tensor(x-mu).float()
 
     
     #Random Initilization of parameters in need of training
-    W = torch.tensor(np.random.normal(0,1,[d,m]))
-    sig = torch.tensor(exp(np.random.normal(0,0.1,[1])))
+    W = torch.tensor(np.random.normal(0,0.1,[d,m])).float()
+    sig = torch.tensor(exp(np.random.normal(0,0.1,[1]))).float()
     
-    #Initialize matrix M
-    #M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
+
 
     # ite iterations of the EM algorithm
     for i in range(0,ite):
         
         #Setting stuff up
-        M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m)
-        Minv = torch.inverse(M)
+        M = torch.matmul(torch.transpose(W,0,1),W)+sig*torch.eye(m).float()
+        Minv = torch.inverse(M).float()
 
         
         #E Step
-        Ez =  torch.matmul(torch.matmul(Minv,torch.transpose(W,0,1)),torch.transpose(xd,0,1))
-        Ez = Ez.view(100,2,1)
-        Ezt = Ez.view(100,1,2)
+        Ez =  torch.matmul(torch.matmul(Minv,torch.transpose(W,0,1)).double(),torch.transpose(xd,0,1).double()).float()
+        Ez = Ez.view(n,m,1)
+        Ezt = Ez.view(n,1,m)
         Ezz = sig*Minv + torch.matmul(Ez,Ezt)
         
         #M Step
[…]

### 2021-08-28 main.py
Deleted old main.py

@@ -1,224 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on 22/01/21
-
-@author: Cedric Beaulac 
-
-LVM-MEGA
-"""
-
-
-####################################
-# Import the good stuff (bunch of libraries)
-####################################
-from __future__ import print_function
-
-import numpy as np
-from numpy import *
-
-import pandas as pd
-
-import torch
-import torch.nn as nn
-import torch.nn.init as init
-import torch.nn.functional as F
-import torch.optim as optim
-from torch.optim.lr_scheduler import StepLR
-import torch.utils.data
-from torch.autograd import Variable
-
-import argparse
-
-import torchvision
-from torchvision import transforms
-from torchvision import datasets, transforms
-from torchvision.utils import save_image
-
-from scipy import stats, integrate
-
-from sklearn.ensemble import VotingClassifier
-from sklearn.metrics import accuracy_score
-from sklearn.preprocessing import LabelEncoder
-from sklearn.preprocessing import OneHotEncoder
-
-import os
-
-import cv2
-
-import seaborn as sns
-
-import matplotlib.pyplot as plt
-
-
-####################################
-# Import my libraries
-####################################
-
-from utility import *
-from NNutility import *
-
-
-
-####################################
-# Generate data sets
-####################################
-
-
-n=5000
-pi=[0.3,0.3,0.4]
-mu=[[0,10],[-5,0],[5,2]]
-std=[[1,2],[0.5,0.5],[2,2]]
-
-sample = GMM_Gen2(n,pi,mu,std)
-
-plt.scatter(sample[:,0].numpy(),sample[:,1].numpy())
-
-
-###########################################
-# Fixing arguments
-###########################################
-
-
-ObsDim = shape(mu)[1]
-BSize = 500
-NB = int(n/BSize)
-
-parser = argparse.ArgumentParser(description='LVM MEGA')
-parser.add_argument('--batch-size', type=int, default=BSize, metavar='N',
-                        help='input batch size for training (default: 64)')
-parser.add_argument('--Number-batch', type=int, default=NB, metavar='N',
-                        help='input batch size for testing (default: 1000)')
-parser.add_argument('--test-batch-size', type=int, default=1000, metavar='N',
[…]
